---
title: "R Notebook"
output: html_notebook
---

Visualizations using C.D.'s NBA data.

```{r}
require(tidyverse)
require(plotly)
require(RColorBrewer)

theme_set(theme_bw())
#readRDS(file = "GE.Rdata")
source("nbacoaches-3.R")

```

Start with some summary statistics type plots.  For example, how long has the current coach been with the team (current being 2020 season).  The columns will be color-coded by the coaches win percentage with the team.

```{r}
pDF <- coach_df %>% filter(season == 2020) %>% 
  arrange(desc(seasons.w.franch)) %>%
  mutate(coach = factor(coach, levels = coach)) %>%
  mutate(win_pct = reg.franch.W/reg.franch.G * 100)
ggplot(pDF, aes(coach, seasons.w.franch)) + geom_col(aes(fill = win_pct)) +
  labs(x = "Coach", 
       y = "Seasons with Franchise", 
       title = "As of 2020",
       fill = "Win %") +
  theme(axis.text.x = element_text(angle = 90))
```

How about average coach tenure by year?  

```{r}
pDF <- coach_df %>% group_by(season) %>%
  summarise(Mean_Tenure = mean(seasons.w.franch),
            Mean_Playoff = sum(playoff.career.G, na.rm = TRUE) / n()) 
ggplot(pDF, aes(season, Mean_Tenure)) + geom_col(aes(fill = Mean_Playoff)) +
  labs(x = "Season", 
       y = "Mean Tenure", 
       title = "Coaching Tenure in the NBA",
       fill = "Mean Number of\n Playoff Games") 
```

How about coaches winning percentages through time.  This won't be labelled by coach, but rather a series of lines extending across seasons with each season's WP record.  With all the coaches, it should be a cloud.

```{r}
pDF <- coach_df %>% group_by(coach, season) %>% 
  summarize(WP = reg.current.W / reg.current.G * 100)
ggplot(pDF, aes(season, WP, group = coach)) + geom_path() +
  labs(x = "Season", y = "Win Pct w/Coach")
```
One last coach plot: winning percentage when fired mid-season.

```{r}
pDF <- coach_df %>% filter(replaced == "midseason") %>% 
  mutate(WP = round(reg.current.W / reg.current.G * 100, 2)) %>%
  mutate("Seasons with Franchise" = `seasons.w.franch`)
         
ggplot(pDF, aes(season, WP)) + 
  geom_point(aes(size = reg.current.G, color = seasons.w.franch)) +
  scale_color_distiller(palette = "Spectral", direction = -1) +
  labs(x = "Season", y = "Win Pct When Fired", 
       title = "Midseason Firings",
       size = "Games\nCoached",
       color = "Seasons with\nFranchise")
```

Let's recreate the previous plot using plotly

```{r}
fig <- plot_ly(data = pDF) %>%
  add_markers(mode = "markers",
              type = "scatter",
              x = ~season,
              y = ~WP,
              color = ~`Seasons with Franchise`,
              colors = rev(brewer.pal(n = 11,
                                      name = "Spectral")),
              size = ~reg.current.G,
              text = ~paste("Coach: ", coach, 
                            '<br>Team:', team,
                            '<br>Games:', reg.current.G,
                            '<br>Seasons:', `Seasons with Franchise`,
                            '<br>Win %:', WP)) %>%
  layout(
    title = "Midseason Firings",
    xaxis = list(title = "Season"),
    yaxis = list(title = "Winning Pct When Fired"),
    showlegend = FALSE
  ) %>% colorbar(title = list(text = "Seasons with\nFranchise"))

fig 
```

Compare WP pre and post midseason firings by year.

```{r}
pDF1 <- coach_df %>% filter(replaced == "midseason") %>%
  group_by(season) %>%
  summarize(Before = sum(reg.current.W) / sum(reg.current.G) * 100, season = season) %>%
  distinct()
pDF2 <- coach_df %>% filter(hired == "midseason") %>%
  group_by(season) %>%
  summarize(After = sum(reg.current.W) / sum(reg.current.G) * 100, season = season) %>%
  distinct()
pDF <- full_join(pDF1, pDF2, by = "season") %>% 
  gather(key = "Which", value = "WP", -season)

ggplot(pDF, aes(season, WP, group = Which)) + geom_col(aes(fill = Which), position = "dodge")
```

```{r}
pDF <- full_join(pDF1, pDF2, by = "season") %>% 
  mutate(Change = After - Before)

m1 <- lm(Change ~ season, pDF)

pDF$fitted <- fitted(m1)
midseason <- pDF

#ggplot(pDF, aes(season, Change)) + geom_point() + geom_line(aes(y = fitted))

ggplot(pDF, aes(season, Change)) + geom_point() + geom_smooth(aes(y = fitted), method = "lm")
```

Compare WP before and after an offseason firing.

```{r}
pDF1 <- coach_df %>% filter(replaced == "offseason" & season < 2020) %>%
  group_by(season) %>%
  summarize(Before = sum(reg.current.W) / sum(reg.current.G) * 100, season = season) %>%
  distinct()
pDF2 <- coach_df %>% filter(hired == "offseason") %>%
  group_by(season) %>%
  summarize(After = sum(reg.current.W) / sum(reg.current.G) * 100, season = season - 1) %>%
  distinct()
pDF <- full_join(pDF1, pDF2, by = "season") %>% 
  mutate(Change = After - Before)

m2 <- lm(Change ~ season, pDF)

pDF$fitted <- fitted(m2)
offseason <- pDF

#ggplot(pDF, aes(season, Change)) + geom_point() + geom_line(aes(y = fitted))

ggplot(pDF, aes(season, Change)) + geom_point() + geom_smooth(aes(y = fitted), method = "lm")
```

Is it better to fire coach mid-season or offseason?

```{r}
wpchange <- tibble(season = offseason$season, 
                   offseason = offseason$Change, 
                   midseason = midseason$Change[-19]) %>%
  summarise(wpoff = mean(offseason), wpmid = mean(midseason))


```

Start looking at effect on lead WS player before/after offseason coaching change.

```{r}
# join player_df and coach_df by season and team

full_df <- left_join(player_df, coach_df, by = c("Season" = "season", "Team" = "team"))
```

```{r}
# filter coach_df by replaced and hired == offseason
coach_change <- full_df %>% filter(replaced == "offseason" | hired == "offseason")
```

Group by season and team and create a dataframe with just the top player

```{r}
top_player_pre <- coach_change %>% group_by(Season, Team) %>% slice_max(WS, n = 1) %>% mutate(TopWS = TRUE)
```

Now find cases where the player remained on the team in subsequent year and pull their stats.

```{r}
player_df_back <- player_df %>% mutate(Season = Season - 1)
player_pre_post <- left_join(top_player_pre, player_df_back, by = c("Player", "Season", "Team")) %>% arrange(Player)
```

```{r}
Player_change <- player_pre_post %>% ungroup() %>%
                transmute(Season = Season,
                          BPM = BPM.y - BPM.x,
                          WS = WS.y - WS.x,
                          VORP = VORP.y - VORP.x) 

Yearly_change <- Player_change %>% group_by(Season) %>% 
  summarize(BPM = mean(BPM, na.rm = TRUE), 
            VORP = mean(VORP, na.rm = TRUE), 
            WS = mean(WS, na.rm = TRUE))
```

```{r}
ggplot(Player_change, aes(VORP)) + geom_histogram(bins = 12)
```

```{r}
# Test if change in BPM is significant
t.test(Player_change$BPM)
t.test(Player_change$WS)
t.test(Player_change$VORP)

```

```{r}
ggplot(Yearly_change, aes(Season, BPM)) + geom_line() + geom_smooth(method = "lm")
ggplot(Yearly_change, aes(Season, VORP)) + geom_line() + geom_smooth(method = "lm")
ggplot(Yearly_change, aes(Season, WS)) + geom_line() + geom_smooth(method = "lm")

```

```{r}
#creating dataframe of mideseason offseason and change
season <- rbind(data.frame(season = "midseason", midseason),
                data.frame(season = "offseason", offseason)) %>%
  select(season, Change)
  
coach_change_model <- lm(Change ~ season, data = season)
summary(coach_change_model)
```

```{r}
midseason_coach_changes <- coach_df %>%
  filter(replaced == "midseason" | hired =="midseason")

```
```{r}
offseason_coach_changes <- coach_df %>%
  filter(replaced == "offseason" | hired =="offseason") %>% 
  mutate(transition = case_when(!is.na(hired) & !is.na(replaced) ~ "in_and_out",
                                !is.na(hired) & is.na(replaced) ~ "incoming",
                                is.na(hired) & !is.na(replaced) ~ "outgoing"),
         trans_year = case_when())

offseason_coach_changes %>%
  group_by(transition) %>%
  summarise(reg.current.wp = sum(reg.current.W)/sum(reg.current.G))

save(offseason_coach_changes, file = "off_season_coach_changes.RData")
load("off_season_coach_changes.RData")
  
```


